---
title: "How do I ... with the tidyverse?"
author: "Jamie Lendrum"
tags: [r, tidyverse]
---

Below is a searchable table containing base R functions and their equivalent tidyverse functions. The purpose of this page is to allow people to either:

* Search for the appropriate tidyverse function given a base R function, or to
* Search for an equivalent base R function when using the tidyverse is not possible/appropriate.

I'm not a fountain of all knowledge, so if you would like to contribute to this table please submit a pull request on the following file: https://github.com/jl5000/r-house/blob/master/content/page/base-r-tidyverse-functions.Rmd (or open an issue).

Happy hunting!


```{r include=FALSE}
df <- tibble::tribble(
  ~`Tidyverse package`,      ~Function, ~`Base R Function`,
             "dplyr",       "select",   "df[], subset",
             "dplyr",        "slice",   "df[], subset",
             "dplyr",       "filter",   "df[], subset",
             "dplyr",         "pull",    "df$, df[[]]",
             "dplyr",      "arrange",  "df[order(df)]",
            "tibble",      "glimpse",            "str",
             "dplyr",      "if_else",         "ifelse",
             "dplyr",     "distinct",         "unique",
             "dplyr",    "left_join",          "merge",
             "dplyr",   "right_join",          "merge",
             "dplyr",   "inner_join",          "merge",
             "dplyr",    "full_join",          "merge",
             "dplyr",    "summarize",      "aggregate",
             "dplyr",    "bind_rows",          "rbind",
             "dplyr",    "bind_cols",          "cbind",
             "dplyr",       "mutate",      "transform",
           "stringr",        "str_c",         "paste0",
           "stringr",   "str_length",          "nchar",
           "stringr",   "str_detect",          "grepl",
           "stringr",   "str_subset",           "grep",
           "stringr",    "str_split",       "strsplit",
           "stringr", "str_to_upper",        "toupper",
           "stringr", "str_to_lower",        "tolower",
             "readr",   "read_delim",     "read.table",
             "readr",     "read_csv",       "read.csv",
             "readr",     "read_tsv",     "read.delim",
             "tidyr",      "drop_na",        "na.omit",
             "tidyr",         "nest",             "by",
             "purrr",          "map",         "lapply",            
             "purrr",        "map_*",         "vapply",                            
             "purrr",   "map2, pmap",         "mapply",     
           "ggplot2",       "ggplot",           "plot"
  )
```

```{r echo=FALSE}
tbl <- DT::datatable(df, 
                     rownames = FALSE,
                     options = list(pageLength = 9,
                                   order = list(list(0, 'asc'), list(1, 'asc'))))
widgetframe::frameWidget(tbl)
```